---
title: "cleaning_and_visuals"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#store string containing all required packages
my_packages <- c('rtweet', 'ggplot2', 'igraph', 'RColorBrewer', 'readr', 'bipartite')
```

Figure out which of these packages is already installed
``` {r, message=F, warning=F}
# Store all installed packages
ya_installed <- library()$results[,1]

# Check whether required packages are already installed and grab only those that still need installation
need_install<-my_packages[!(my_packages %in% ya_installed)]

#install required packages
lapply({need_install}, install.packages, character.only = TRUE)

```


``` {r, message=F, warning=F}
# Store all installed packages
ya_loaded <- (.packages())

# Check whether required packages are already installed and grab only those that still need installation
need_load<-my_packages[!(my_packages %in% ya_loaded)]

# Load required packages
lapply(need_load, require, character.only = TRUE)

#also load the readr library
library(readr)
library(bipartite)
library(dplyr)
library(tidyr)
library(lubridate)
library(ggplot2)
library(igraph)
library(reshape2)

```

```{r}
# Load data (all columns read as characters initially)
data <- read_csv("org_full_time_series.csv", col_types = cols(.default = col_character()))

data <- data %>% select(-c(228,229,230,231))

# Convert numeric columns (e.g., UPB, face values, etc.) by removing "$" and ","
numeric_cols <- names(data)[7:ncol(data)]  # assuming numeric data start from column 7
data[numeric_cols] <- data[numeric_cols] %>% 
  mutate(across(everything(), ~parse_number(.)))

# Merge duplicate DC columns if present
if("District Of Columbia_Aggregate_UPB" %in% names(data)) {
  data <- data %>%
    mutate(`District of Columbia_Aggregate_UPB` = coalesce(`District of Columbia_Aggregate_UPB`, `District Of Columbia_Aggregate_UPB`),
           `District of Columbia_Percent_UPB` = coalesce(`District of Columbia_Percent_UPB`, `District Of Columbia_Percent_UPB`),
           `District of Columbia_Number_of_Loans` = coalesce(`District of Columbia_Number_of_Loans`, `District Of Columbia_Number_of_Loans`),
           `District of Columbia_Percent_Loans` = coalesce(`District of Columbia_Percent_Loans`, `District Of Columbia_Percent_Loans`)) %>%
    select(-starts_with("District Of Columbia_"))
}

# Convert date columns to Date type
data <- data %>%
  mutate(`Purchase Date` = as.Date(`Purchase Date`),
         `Sale Date` = as.Date(`Sale Date`))
         
# Create HoldingDays variable (difference between Sale Date and Purchase Date)
data <- data %>% mutate(HoldingDays = as.numeric(`Sale Date` - `Purchase Date`))

# Peek at the data structure
glimpse(data)

```

```{r}
data_filtered <- data %>%
  filter(year(`Purchase Date`) %in% c(2009:2012, 2019:2022))
```

```{r}
# Reshape state-level UPB columns to long format
state_upb_long <- data_filtered %>%
  select(CUSIP, `Purchase Date`, ends_with("_Aggregate_UPB")) %>%
  pivot_longer(
    cols = ends_with("_Aggregate_UPB"),
    names_to = "State",
    names_pattern = "(.*)_Aggregate_UPB",
    values_to = "UPB"
  ) %>%
  mutate(UPB = as.numeric(UPB),
         Year = year(`Purchase Date`)) %>%
  filter(!is.na(Year))

# Compute total UPB per CUSIP (summing across all states)
cusip_totals <- state_upb_long %>%
  group_by(CUSIP) %>%
  summarize(Total_UPB = sum(UPB, na.rm = TRUE), .groups = "drop")

# Merge total UPB back and calculate percentage per state
cusip_composition <- state_upb_long %>%
  left_join(cusip_totals, by = "CUSIP") %>%
  mutate(Percent_State_UPB = if_else(Total_UPB > 0, 100 * UPB / Total_UPB, 0))

```

```{r}
# For each CUSIP, compute the maximum state percentage
cusip_exposure <- cusip_composition %>%
  group_by(CUSIP) %>%
  summarize(Max_Percent = max(Percent_State_UPB, na.rm = TRUE), .groups = "drop")

# Merge the exposure data with the holding period information
cusip_exposure <- cusip_exposure %>%
  left_join(data_filtered %>% select(CUSIP, HoldingDays), by = "CUSIP") %>%
  mutate(Exposure = if_else(Max_Percent > 50, "Concentrated", "Diverse"))

# Check summary
summary(cusip_exposure)

```

```{r}
ggplot(cusip_exposure, aes(x = Exposure, y = HoldingDays, fill = Exposure)) +
  geom_boxplot() +
  labs(title = "Holding Period by UPB Exposure Concentration",
       x = "Exposure Type",
       y = "Holding Period (days)") +
  theme_minimal()
```

```{r}

#Build a small long‐format df with only positive %UPB
mbs_web_df <- cusip_composition%>%
  filter(Percent_State_UPB > 0) %>%
  select(State, CUSIP, Percent_State_UPB)

mbs_web_df$grouping_factor <- "gf"

#Use acast() to pivot to a matrix: rows = States, cols = CUSIPs
incidence_mat <- acast(
  mbs_web_df,
  State   ~ CUSIP,
  value.var  = "Percent_State_UPB",
  fun.aggregate = sum,
  fill         = 0
)

#Coerce to "web" so plotweb() does CA under the hood
class(incidence_mat) <- c("web", class(incidence_mat))


cols1 <- c(
  '#8dd3c7','#ffffb3','#bebada','#fb8072',
  '#80b1d3','#fdb462','#b3de69','#fccde5',
  '#d9d9d9','#bc80bd','#ccebc5','#ffed6f'
)

#Plot the bipartite graph with correspondence‐analysis layout
plotweb(
  incidence_mat,        # our State×CUSIP %UPB matrix
  method            = "cca",   # same as the women example
  labsize           = 1,
  text.rot          = 90,
  col.interaction   = cols1,
  bor.col.interaction = cols1
)

```

In the chunk above, I first used acast() to build incidence_mat, a matrix with rows = States and columns = CUSIPs, whose entries are the percent of unpaid principal balance (UPB) contributed by each state to each security. I then coerced that matrix to class "web" and called, plotweb( incidence_mat, method="cca", … )
which runs a Correspondence Analysis (CA) on these weighted link values. 

Because we pass the actual % UPB into CA, securities that draw their balances from very similar state mixes are placed close together in the ordination, and likewise for states that have similar exposure profiles.

From the plot we can see major hubs like California and New York are noticiable by the extremely dense fans of lines radiating upward at their tick marks showing how they dominate many securities. Sub‑clusters of states (e.g. Midwest vs. Pacific vs. Northeast) whose line‑bundles overlap in similar regions, indicating geographic or market‐style groupings.Outlier territories (Guam, Puerto Rico) with only a few thin lines off to the side, reflecting their minimal footprint in the Fed’s MBS holdings.

```{r}
# Concentrated weighted web
conc_ids <- cusip_exposure %>% filter(Exposure=="Concentrated") %>% pull(CUSIP)
inc_mat_conc <- acast(
  cusip_composition     %>% filter(CUSIP %in% conc_ids & Percent_State_UPB>0),
  State ~ CUSIP,
  value.var = "Percent_State_UPB",
  fun.aggregate = sum,
  fill = 0
)
class(inc_mat_conc) <- c("web","matrix")
```

```{r}
library(bipartite)

# Concentrated
net_conc_w <- networklevel(
  inc_mat_conc,
  index    = c("connectance","nestedness","robustness"),
  level    = "both",
  weighted = TRUE
)
grp_conc_w <- grouplevel(
  inc_mat_conc,
  level    = "both",
  weighted = TRUE
)
sp_conc_w  <- specieslevel(
  inc_mat_conc,
  level    = "both",
  weighted = TRUE
)
```

```{r}
library(igraph)

# State–state for concentrated, weighted by sum of %UPB overlaps
g_conc_w <- graph_from_adjacency_matrix(
  inc_mat_conc %*% t(inc_mat_conc),
  mode="undirected", weighted=TRUE, diag=FALSE
)
# Now compute weighted density, clustering, etc.
edge_density(g_conc_w)                      # weighted density ≈ connectance
transitivity(g_conc_w, type="global")       # weighted clustering
vcount(g_conc_w)                        # number of states                # number of 
diameter(g_conc_w, weights=NA)
```

Calculate statistics for the diverse web. 
```{r}
# Diverse weighted web
div_ids  <- cusip_exposure %>% filter(Exposure=="Diverse")    %>% pull(CUSIP)
inc_mat_div  <- acast(
  cusip_composition     %>% filter(CUSIP %in% div_ids & Percent_State_UPB>0),
  State ~ CUSIP,
  value.var = "Percent_State_UPB",
  fun.aggregate = sum,
  fill = 0
)
class(inc_mat_div) <- c("web","matrix")

```

```{r}
library(bipartite)

# Concentrated
net_div_w <- networklevel(
  inc_mat_div,
  index    = c("connectance","nestedness","robustness"),
  level    = "both",
  weighted = TRUE
)
grp_div_w <- grouplevel(
  inc_mat_div,
  level    = "both",
  weighted = TRUE
)
sp_div_w  <- specieslevel(
  inc_mat_div,
  level    = "both",
  weighted = TRUE
)
```

```{r}
library(igraph)

# State–state for concentrated, weighted by sum of %UPB overlaps
g_div_w <- graph_from_adjacency_matrix(
  inc_mat_div %*% t(inc_mat_div),
  mode="undirected", weighted=TRUE, diag=FALSE
)
# Now compute weighted density, clustering, etc.
edge_density(g_div_w)                      # weighted density ≈ connectance
transitivity(g_div_w, type="global")       # weighted clustering
vcount(g_div_w)                        # number of states                # number of 
diameter(g_div_w, weights=NA)
```

```{r}
library(dplyr)
library(reshape2)
library(bipartite)

# 12‐color palette
cols1 <- c(
  '#8dd3c7','#ffffb3','#bebada','#fb8072',
  '#80b1d3','#fdb462','#b3de69','#fccde5',
  '#d9d9d9','#bc80bd','#ccebc5','#ffed6f'
)

# --- A) Concentrated web ---
incidence_mat_conc <- cusip_composition %>%
  # bring in the Exposure label
  inner_join(
    cusip_exposure %>% select(CUSIP, Exposure),
    by = "CUSIP"
  ) %>%
  filter(Exposure == "Concentrated", Percent_State_UPB > 0) %>%
  acast(
    State ~ CUSIP,
    value.var     = "Percent_State_UPB",
    fun.aggregate = sum,
    fill          = 0
  )

class(incidence_mat_conc) <- c("web", class(incidence_mat_conc))

plotweb(
  incidence_mat_conc,
  method            = "cca",
  labsize           = 1,
  text.rot          = 90,
  col.interaction   = cols1,
  bor.col.interaction = cols1,
)
title("Concentrated CUSIPs (Weighted %UPB)")
```


```{r}
library(dplyr)
library(bipartite)

set.seed(123)

# 1) Pick a random subset of States and CUSIPs
all_states  <- rownames(inc_mat_div)
all_cusips  <- colnames(inc_mat_div)

# here I sample 25 states and 80 CUSIPs (tune as needed)
sample_states_div <- sample(all_states, 53)
sample_cusips_div <- sample(all_cusips, 100)

# 2) Subset the full matrix
sub_div_mat <- inc_mat_div[sample_states_div, sample_cusips_div, drop = FALSE]

# 3) Coerce to "web" class
class(sub_div_mat) <- c("web", class(sub_div_mat))

# 4) Define the same 12‑color palette
cols1 <- c(
  '#8dd3c7','#ffffb3','#bebada','#fb8072',
  '#80b1d3','#fdb462','#b3de69','#fccde5',
  '#d9d9d9','#bc80bd','#ccebc5','#ffed6f'
)

# 5) Plot with CA layout on the sampled submatrix
plotweb(
  sub_div_mat,
  method            = "cca",
  labsize           = 1,
  text.rot          = 90,
  col.interaction   = cols1,
  bor.col.interaction = cols1
)
title("Diverse CUSIPs (Weighted %UPB) — Sampled Subnetwork")

```



